Hierarchical Time Series Forecasting in Emergency Medical Services

### 2022-11-17 manuscript/03-experiment.qmd
Working on sec 3

We are interested in generating forecasts to inform the planning horizon of $ph= 42$ days, required by planners in the ambulance services trust. The forecast horizon in this study is $fh = 2 \times ph$ days ahead (2*42 days planning horizon). This is because the planning is generally frozen for $ph$ days and considering a forecast horizon of $ph$ days might not be helpful for planning. While forecasts are generated for $2 \times ph$ days ahead, performance evaluation is only assessed based on the last $ph$ days and not the $2 \times ph$ days. The forecasts are produced for the holdout of 365 days using time series cross-validation [@hyndman2021forecasting].

In the following section, we discuss the dataset, describe the forecasting methods used to generate base forecasts and present the point and probabilistic accuracy measures.

@@

The dataset used in this study is from a major ambulance service trust in Great Britain. It contains information relating to the daily number of attended incidents from 1 October 2015 to 31 July 2019, disaggregated by nature of incidents, priority, the health board managing the service and the control area (or region). @fig-hierarchy depicts both the hierarchical and grouped structure of the data. @fig-hierarchy-1 illustrates the nested hierarchical structure based on control area and health board and @fig-hierarchy-2 shows the grouped structure by priority and the nature of incident.

```{r}
#| label: fig-hierarchy
#| out.width: "60%"
#| fig-cap: "The hierarchical and grouped structure of attended incidents (ambulance demand)."
#| fig-subcap:
#|   - "Hierarchical structure: Attended incidents in all country is disaggregated into 3 control areas /regions and then into 7 different healthboard at the bottom"
#|   - "Grouped structure:  Incidents could be grouped into priority (i.e. Red, Amber & Green) and the nature of attended incident (i.e. there are 35 different nature of incidents including chest pain , breathing problems, heart attack, stroke, and so on). The symbol * refers to the crossed attributes between hierarchical and grouped levels."
#| layout: [[50],[-2], [38]]

data <- data.frame(
  level1 = "Total",
  level2 = c("Central & West", "Central & West", "Central & West",
             "North", "South & East", "South & East", "South & East"),
  level3 = c("HD", "SB", "PO", "BC", "CV", "CT", "AB")
)
# transform it to a edge list!
edges_level1_2 <- data %>%
  select(level1, level2) %>%
  unique() %>%
  rename(from = level1, to = level2)
edges_level2_3 <- data %>%
  select(level2, level3) %>%
  unique() %>%
  rename(from = level2, to = level3)
edge_list <- rbind(edges_level1_2, edges_level2_3)

mygraph <- igraph::graph_from_data_frame(edge_list)
ggraph::ggraph(mygraph, layout = "dendrogram", circular = FALSE) +
  ggraph::geom_edge_diagonal() +
  ggraph::geom_node_point(color = "#dddddd", size = 10) +
  ggraph::geom_node_text(
    aes(label = c("All country",
                  "Central & West", "North", "South & East",
                  "HD", "SB", "PO", "BC", "CV", "CT", "AB"))
  ) +
  theme_void()

@@

```{r}
#| label: tbl-hierarchy
#| tbl-cap: "Number of time series in each level for the hierarchical & grouped structure of attended incidents"
agg_level <- tibble::tribble(
  ~Level, ~`Number of series`,
  "All country", 1,
  "Control", 3,
  "Health board", 7,
  "Priority", 3,
  "Priority * Control", 9,
  "Priority * Health board", 21,
  "Nature of incident", 35,
  "Nature of incident * Control", 105,
  "Nature of incident * Health board", 245,
  "Priority * Nature of incident", 104,
  "Control * Priority * Nature of incident", 306,
  "Control * Health board * Priority * Nature of incident (Bottom level)", 691,
  "Total", 1530
)
knitr::kable(agg_level, booktabs = TRUE, position = "left") %>%
  kable_classic(full_width = FALSE)
```

Given the total number of time series, direct visual analysis is infeasible. Therefore, we first compute features of all 1530 time series [@m3pca] and display the strength of trend and weekly seasonality strength in @fig-feature. Each point represents one time series with the strength of trend in x-axis and the strength of seasonality in y-axis. It is clear that there are some series showing strong trends and/or seasonality, corresponding to series at the higher levels of the hierarchy. The majority of series show low trend and seasonality. These are time series belonging to the bottom series, series related to the nature of incidents for a given control, health board and priority level. Bottom series are dominated by noise with little or no systematic patterns.

@@

In addition to displaying the trend and seasonality features, we also visualize few time series at various levels of the aggregation. @fig-dataviz2 reveals different information such as trend, seasonality and noise. For example, some series depict seasonality and trend, whereas some other series report low volume of attended incidents and entropy, making them more volatile and difficult to forecast. At the level on nature of incidents combined with categories of other levels, there are many series that contain zeros with low counts. As such, the data set represents a diverse set of daily time series patterns.

```{r}
#| label: fig-dataviz
#| eval: false
#| include: false
#| dependson: "fig-feature"
#| out.width: "70%"
#| fig-cap: "Time series of attended incidents at various levels."
#| fig-subcap:
#|   - "Whole country"
#|   - "By control area"
#|   - "By health board and control area"
#|   - "By health board and priority"
#|   - "By nature of incident and health board"
#| layout: [[44,-2, 44], [44,-2, 44], [100]]
p_total <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  autoplot() +
  labs(x = "Date", y = "Incidents") +
  ggthemes::theme_few()

p_total_control <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>%
  select(-nature, -category) %>%
  group_by(date, region) %>%
  summarise(incident = sum(incident), .groups = "drop") %>%
  ggplot(aes(x = date, y = incident, color = factor(region))) +
  geom_line() +
  labs(x = "Date", y = "Incidents", color = "Control") +
  ggthemes::scale_color_colorblind() +
  ggthemes::theme_few() +
  theme(legend.position = "bottom")

p_control <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>%
  select(-nature, -category) %>%
  ggplot(aes(x = date, y = incident, color = factor(lhb))) +
  geom_line() +
  facet_wrap(vars(factor(region)), scales = "free_y") +
  labs(x = "Date", y = "Incidents", color = "Health board") +
  ggthemes::scale_color_colorblind() +
  ggthemes::theme_few() +
  theme(legend.position = "bottom")

p_priority <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & !is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>%
  select(-nature, -region) %>%
  ggplot(aes(x = date, y = incident, color = factor(lhb))) +
  geom_line() +
  facet_wrap(vars(factor(category)), scales = "free_y") +
  labs(x = "Date", y = "Incidents", color = "Health board") +
  ggthemes::scale_color_colorblind() +
  ggthemes::theme_few() +
  theme(legend.position = "bottom")

selected_nature <- c("CHESTPAIN", "STROKECVA", "BREATHING", "ABDOMINAL")

p_nature_healthboard <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & !is_aggregated(category) & !is_aggregated(nature)) %>%
  as_tibble() %>%
  mutate(nature = as.character(nature)) %>%
  filter(nature %in% selected_nature) %>%
  group_by(date, nature, lhb) %>%
  summarise(incident = sum(incident), .groups = "drop") %>%
  ggplot(aes(x = date, y = incident, color = factor(lhb))) +
  geom_line() +
  facet_wrap(vars(factor(nature)), scales = "free_y") +
  labs(x = "Date", y = "Incidents", color = "Health board") +
  ggthemes::scale_color_colorblind() +
  ggthemes::theme_few() +
  theme(legend.position = "none")

p_total
p_total_control
p_control
p_priority
p_nature_healthboard
```

```{r}
#| label: fig-dataviz2
#| dependson: "fig-feature"
#| out.width: "100%"
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "Time series of attended incidents at various levels."
no_x_axis <- theme(axis.title.x=element_blank(),
                   axis.text.x=element_blank(),
                   axis.ticks.x=element_blank())
p_total <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  autoplot(incident) +
  labs(x = "", y = "Incidents") +
  no_x_axis

@@

p_board <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>%
  select(-nature, -category) %>%
  ggplot(aes(x = date, y = incident, color = factor(lhb))) +
  geom_line() +
  #facet_wrap(vars(factor(region)), scales = "free_y") +
  labs(y = "Incidents", color = "Health board") +
  no_x_axis

p_priority <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & !is_aggregated(category) & is_aggregated(nature)) %>%
  mutate(
    category = recode(category, RED = "Red", AMB = "Amber", GRE = "Green"),
    category = factor(category, levels=c("Red","Amber","Green"))
  ) |>
  as_tibble() %>%
  select(-nature, -region) %>%
  ggplot(aes(x = date, y = incident, color = factor(category))) +
  geom_line() +
  scale_color_manual(values = c(Red="#ff3300", Amber = "#E69f00", Green="#009e73")) +
  labs(y = "Incidents", color = "Priority") +
  no_x_axis

@@

Given the presence of various significant patterns in the past attended incidents, we consider three different forecasting models to generate the base forecasts. Once the base forecasts are produced, hierarchical and grouped time series methods are used to reconcile them across the all levels. We briefly discussed forecasting models in the following sections, and the hierarchical forecasting methods are discussed in @sec-htc.

@@

**Exponential Smoothing State Space model (ETS):** ETS models [@hyndman2021forecasting] can combine trend, seasonality and error components in a time series through various forms such as additive, multiplicative or mixed. The trend component can be none ("N"), Additive ("A"), damped ("Ad") or multiplicative ("M"). The seasonality can be none ("N"), Additive ("A"), or multiplicative ("M"). The error term can also be additive ("A") or multiplicative ("M"). To forecast the attended incidents at each level, we use the `ETS()` function in the `fable` package [@fable2022] in R. To identify the best model for a given time series, the ETS function uses the corrected Akaike’s Information Criterion (AICc).

Despite the popularity and the relevance of automatic `ETS` in this study, however it may produce forecast distributions that are non-integer and include negative values, despite the number of attended incidents being integer and non-negative. When using `ETS`, a time series transformation approach could be used to generate strictly positive forecasts, although forecast distributions will still be non-integer. An alternative is to use  forecasting models that produce integer, non-negative forecasts. in the following section we present Generalized Linear Models (GLMs) and Poisson Regression to produce count base forecasts.

>>>>

We are interested in generating forecasts to inform the planning horizon of 42 days, required by planners in the ambulance service. The forecast horizon in this study is $2 \times 42 = 84$ days ahead, because the planning is generally frozen for 42 days and so forecasts of the next 42 days is not particularly helpful for planning. While forecasts are generated for 84 days ahead, performance evaluation is only assessed based on the last 42 days and not the whole forecast period. The forecasts are produced for various training and test sets using time series cross-validation [@hyndman2021forecasting].

In the following section, we discuss the dataset, describe the forecasting methods used to generate base forecasts, and present the point and probabilistic accuracy measures.

@@

The dataset used in this study is from a major ambulance service in Great Britain. It contains information relating to the daily number of attended incidents from 1 October 2015 to 31 July 2019, disaggregated by nature of incidents, priority, the health board managing the service and the control area (or region). @fig-hierarchy depicts both the hierarchical and grouped structure of the data. @fig-hierarchy-1 illustrates the nested hierarchical structure based on control area and health board and @fig-hierarchy-2 shows the grouped structure by priority and the nature of incident.

```{r}
#| label: fig-hierarchy
#| cache: true
#| out.width: "60%"
#| fig-cap: "The hierarchical and grouped structure of attended incidents (ambulance demand)."
#| fig-subcap:
#| - "Hierarchical structure: Attended incidents in all country is disaggregated into 3 control areas /regions and then into 7 different healthboard at the bottom"
#| - "Grouped structure: Incidents could be grouped into priority (i.e. Red, Amber & Green) and the nature of attended incident (i.e. there are 35 different nature of incidents including chest pain , breathing problems, heart attack, stroke, and so on). The symbol * refers to the crossed attributes between hierarchical and grouped levels."
#| layout: [[50],[-2], [38]]

data <- data.frame(
  level1 = "Total",
  level2 = c(
    "Central & West", "Central & West", "Central & West",
    "North", "South & East", "South & East", "South & East"
  ),
  level3 = c("HD", "SB", "PO", "BC", "CV", "CT", "AB")
)
# transform it to a edge list!
edges_level1_2 <- data %>%
  select(level1, level2) %>%
  unique() %>%
  rename(from = level1, to = level2)
edges_level2_3 <- data %>%
  select(level2, level3) %>%
  unique() %>%
  rename(from = level2, to = level3)
edge_list <- rbind(edges_level1_2, edges_level2_3)

mygraph <- igraph::graph_from_data_frame(edge_list)
ggraph::ggraph(mygraph, layout = "dendrogram", circular = FALSE) +
  ggraph::geom_edge_diagonal() +
  ggraph::geom_node_point(color = "#dddddd", size = 10) +
  ggraph::geom_node_text(
    aes(label = c(
      "All country",
      "Central & West", "North", "South & East",
      "HD", "SB", "PO", "BC", "CV", "CT", "AB"
    ))
  ) +
  theme_void()

@@

```{r}
#| label: tbl-hierarchy
#| cache: true
#| tbl-cap: "Number of time series in each level for the hierarchical & grouped structure of attended incidents"
agg_level <- tibble::tribble(
  ~Level, ~`Number of series`,
  "All country", 1,
  "Control", 3,
  "Health board", 7,
  "Priority", 3,
  "Priority * Control", 9,
  "Priority * Health board", 21,
  "Nature of incident", 35,
  "Nature of incident * Control", 105,
  "Nature of incident * Health board", 245,
  "Priority * Nature of incident", 104,
  "Control * Priority * Nature of incident", 306,
  "Control * Health board * Priority * Nature of incident (Bottom level)", 691,
  "Total", 1530
)
knitr::kable(agg_level, booktabs = TRUE, position = "left") %>%
  kable_classic(full_width = FALSE)
```

Given the total number of time series, direct visual analysis is infeasible. Therefore, we first compute features of all 1530 time series [@m3pca] and display the strength of trend and weekly seasonality strength in @fig-feature. Each point represents one time series with the strength of trend in x-axis and the strength of seasonality in y-axis. Both measures are on a scale of [0,1]. It is clear that there are some series showing strong trends and/or seasonality, corresponding to series at the higher levels of the hierarchy. The majority of series show low trend and seasonality. These are time series belonging to the bottom series, series related to the nature of incidents for a given control, health board and priority level. Bottom series are dominated by noise with little or no systematic patterns.

@@

In addition to displaying the trend and seasonality features, we also visualize a few time series at various levels of the aggregation. @fig-dataviz2 reveals different information such as trend, seasonality and noise. For example, some series depict seasonality and trend, whereas some other series report low volume of attended incidents and entropy, making them more volatile and difficult to forecast. At the level on nature of incidents combined with categories of other levels, there are many series that contain zeros with low counts. As such, the data set represents a diverse set of daily time series patterns.

```{r}
#| label: fig-dataviz2
#| cache: true
#| dependson: "fig-feature"
#| out.width: "100%"
#| fig-width: 8
#| fig-height: 10
#| fig-cap: "Time series of attended incidents at various levels. The panels show data from the whole country, by control area, by health board, by priority level, and by nature of incident. Only four of the 35 nature of incident categories are shown to avoid too much overplotting."
no_x_axis <- theme(
  axis.title.x = element_blank(),
  axis.text.x = element_blank(),
  axis.ticks.x = element_blank()
)
p_total <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  autoplot(incident) +
  labs(x = "", y = "Incidents") +
  no_x_axis

@@

p_board <- incident_gthf %>%
  filter(!is_aggregated(region) & !is_aggregated(lhb) & is_aggregated(category) & is_aggregated(nature)) %>%
  as_tibble() %>%
  select(-nature, -category) %>%
  ggplot(aes(x = date, y = incident, color = factor(lhb))) +
  geom_line() +
  # facet_wrap(vars(factor(region)), scales = "free_y") +
  labs(y = "Incidents", color = "Health board") +
  no_x_axis

p_priority <- incident_gthf %>%
  filter(is_aggregated(region) & is_aggregated(lhb) & !is_aggregated(category) & is_aggregated(nature)) %>%
  mutate(
    category = recode(category, RED = "Red", AMB = "Amber", GRE = "Green"),
    category = factor(category, levels = c("Red", "Amber", "Green"))
  ) |>
  as_tibble() %>%
  select(-nature, -region) %>%
  ggplot(aes(x = date, y = incident, color = factor(category))) +
  geom_line() +
  scale_color_manual(values = c(Red = "#ff3300", Amber = "#E69f00", Green = "#009e73")) +
  labs(y = "Incidents", color = "Priority") +
  no_x_axis

@@

Given the presence of various significant patterns in the past attended incidents, we consider three different forecasting models to generate the base forecasts. Once the base forecasts are produced, hierarchical and grouped time series methods are used to reconcile them across all levels. We briefly discuss forecasting models in the following sections, and the hierarchical forecasting methods are discussed in @sec-htc.

@@

**Exponential Smoothing State Space model (ETS):** ETS models [@hyndman2021forecasting] can combine trend, seasonality and error components in a time series through various forms that can be additive, multiplicative or mixed. The trend component can be none ("N"), Additive ("A") or damped ("Ad"); the seasonality can be none ("N"), Additive ("A"), or multiplicative ("M"); and the error term can be additive ("A") or multiplicative ("M"). To forecast the attended incidents at each level, we use the `ETS()` function in the `fable` package [@fable2022] in R. To identify the best model for a given time series, the `ETS` function uses the corrected Akaike’s Information Criterion (AICc).

Despite the popularity and the relevance of automatic `ETS` in this study, it may produce forecast distributions that are non-integer and include negative values, despite the number of attended incidents being integer and non-negative. When using `ETS`, a time series transformation approach could be used to generate strictly positive forecasts, although forecast distributions will still be non-integer. An alternative is to use forecasting models that produce integer, non-negative forecasts. In the following section we present Generalized Linear Models (GLMs) and Poisson time series regression to produce count base forecasts.
